---
Title: "ProsperEDA"
author: "Priya Ananthram"
date: "18 August 2015"
output: html_document
---
#Univariate plot section
```{r loading_data,cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
loan_data_csv<-read.csv("prosperLoanData.csv")

#The dataset has 113937 observation and 81 variables


library(ggplot2)
library(gridExtra)
library(lubridate)
library(dplyr)
library(tidyr)
library(gridExtra)
loan_data_csv$Loan_year<-year(loan_data_csv$LoanOriginationDate)
loan_data_csv$Loan_month<-month(loan_data_csv$LoanOriginationDate)
loan_data_csv$Loan_closed_Date_month<-month(as.POSIXlt(loan_data_csv$ClosedDate, 
                                                       format="%Y-%m-%d"))
loan_data_csv$Loan_closed_Date_year<-year(as.POSIXlt(loan_data_csv$ClosedDate,
                                                     format="%Y-%m-%d"))
loan_data_csv$Credit_Type<-NA
loan_data_csv[which(loan_data_csv$CreditScoreRangeUpper<=499),'Credit_Type']<-'BAD'
loan_data_csv[which(loan_data_csv$CreditScoreRangeLower>499 ),'Credit_Type']<-'POOR'
loan_data_csv[which(loan_data_csv$CreditScoreRangeLower>599 ),'Credit_Type']<-'FAIR'
loan_data_csv[which(loan_data_csv$CreditScoreRangeLower>659 ),'Credit_Type']<-'GOOD'
loan_data_csv[which(loan_data_csv$CreditScoreRangeLower>779 ),'Credit_Type']<-'VGOOD'

blue.bold.italic.16.text <- element_text(angle = 90, hjust = 1,vjust=1)

```
### Loan Volume by months

I want to first check if there is any variation in loan volumes.Since this data
is US based I want to check if volumes increase or decrease during certain
months i.e during the holiday season,Thanksgiving etc, Since people go on a 
shopping spree during these months defaults on loans may also increase during 
this period is this the case?


```{r loan_volumes_by_month, cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}

most_loans<-ggplot(aes(x=Loan_month),data=loan_data_csv)+
geom_histogram(fill='light blue',binwidth=0.3)+
scale_x_discrete(breaks=seq(1,12),labels=month.abb)+
ggtitle('Loans by month')


most_loans

```

Increase in loan volume during holiday season Oct,Nov,Dec,Jan can be noticed .

###When are most loans defaulted?
```{r loan_volumes_defaulted_by_month, cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}

most_defaulted<-ggplot(aes(x=Loan_closed_Date_month,
                           y=(..count..)/sum(..count..),
                           label=round((..count..)/sum(..count..),2)),
                       data=subset(loan_data_csv ,
                                   LoanStatus %in% c('Defaulted','Chargedoff')))+
geom_histogram(fill='light blue',binwidth=0.3)+
  labs(y="Percent",
       x="Months") +
scale_x_continuous(breaks=seq(1,12),labels=month.abb)+
ggtitle('Loans defaulted by months')

grid.arrange(most_loans,most_defaulted,ncol=1)
```

This distribution is similar to loan volumes by months so defaults seem to
increase during holiday season too


###Loan Volumes over the years 
I want to check if the loan volumes have changed over the years.Prosper started
in 2005 GFC occured in 2008 so did the volumes change over that period.

```{r loan_volumes_by_year ,cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}

ggplot(aes(x=Loan_year),data=loan_data_csv)+
  geom_histogram(fill='light blue',binwidth=0.3)+
  scale_x_continuous(breaks=seq(2005,2014))+
  ggtitle("Loan volumes over the years")

```

It appears loans peaked in 2013 then in the year 2009-2010 dropped off and 
then started picking up again.Note we do not haveall the data for 2014.

It will be interesting to understand why volumes fell in 2009

###Where are the most loans taken
Which state uses Prosper and takes out most loans.

```{r loan_value_by_state, cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}
gg<-ggplot(aes(x=BorrowerState,ymax=max(..count..)),data=loan_data_csv)
gg<-gg+geom_histogram(fill='light blue')
gg<-gg+theme(axis.text.x = blue.bold.italic.16.text)
gg<-gg+stat_bin(aes(y=..count.., label=..count..), geom="text", vjust=-.5,size=3)
gg<-gg+ggtitle("Loans volumes by state")
gg

```

From the graph CA takes out most loans


###How long are the terms of the loan? 
Do people take short term loans or do they take long term loans
```{r loan_value_by_term, cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}
gg<-ggplot(aes(x=Term),data=loan_data_csv)
gg<-gg+geom_histogram(fill='light blue')
gg<-gg+scale_x_discrete(breaks=seq(12,60,12))
gg<-gg+ylim(0,90000)
gg<-gg+ggtitle("Loan terms")
gg
```

Most loans are 36 months 



```{r loan_values,cache=TRUE,echo=FALSE,eval=FALSE, cache.path = 'cache/', fig.path='figure/'}
###How much do people borrow typically?
gg<-ggplot(aes(x=LoanOriginalAmount,ymax=max(..count..)),data=loan_data_csv)
gg<-gg+geom_histogram(fill='light blue',binwidth=300)
gg<-gg+scale_x_continuous(breaks=seq(1000,12000,500),limits=c(1000,12000))
gg<-gg+ggtitle("loan original amounts spread")
gg

summary(loan_data_csv$LoanOriginalAmount)
#Loan amount distribution is slightly positively skewed with median about 6500$.


```
```{r echo=FALSE,eval=FALSE }
###How much percent are loans funded

table(round(loan_data_csv$PercentFunded,2))
#Most loans were fully funded 

```
###How much monthly loan repayment?
```{r loan_repayments,cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}


 summary(loan_data_csv$MonthlyLoanPayment)
 ggplot(aes(x=MonthlyLoanPayment,ymax=max(..count..)),data=loan_data_csv)+
   geom_histogram(fill='lightblue',binwidth=20)+
   scale_x_continuous(breaks=seq(0,1000,50),limits=c(0,1000))+
   #stat_bin(aes(y=..count.., label=..count..), geom="text", vjust=-.5,size=4)+
   ggtitle("Monthly loan payment spread")

```

The loan payment spread is positively skewed with the most common repayment 
about $200

###What interest are people borrowing the loans at?

```{r loan_interst_rates,cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}
 ggplot(aes(x=BorrowerAPR),data=subset(loan_data_csv,!is.na(BorrowerAPR)))+
 geom_histogram(binwidth=0.005,fill='light blue')+
 xlim(c(0.0065,0.52))+
 ggtitle("Borrower APR spread")
 
```

The distribution of borrowerAPR is slightly positively skewed and there is 
spike about 0.35%

###What is the prosperscore of the loans.
Prosper score are a custom risk score built using historical Prosper data 
applicable for loans originated after July 2009. Most loans from the graph seem 
to be between 4-8 prosper score I guess relatively high 
prosper scores should predict a good loan outcome . It would be intersting 
to see if low prosper scores give higher lender yield and vice versa and 
also if prosper scores that are high predict a good loan outcome

```{r loan_prosper_score,cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}
ggplot(aes(x=ProsperScore,ymax=max(..count..)),
       data=subset(loan_data_csv,
                   !is.na(loan_data_csv$ProsperScore)))+

geom_histogram(binwidth=0.4,fill='light blue')+
scale_x_discrete(breaks=seq(1,11,1))
```


###What is the most common reason for the loan
It would be intersting to understand why people are using P2P lending as opposed 
to Banks. For lenders its obvious the yields are higher though risks should be 
higher and due deligence work will be higher for for borrowers the Reasons I
could include turnaround time ,low credit scores ,lower repayments .
```{r loan_reason,cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}

 #table(loan_data_csv$ListingCategory)


blue.bold.italic.16.text <- element_text(angle = 90, hjust = 1,vjust=1)
ggplot(aes(x=ListingCategory..numeric.,ymax=max(..count..)),data=loan_data_csv)+
  geom_histogram(fill='light blue',binwidth=1)+
  scale_x_discrete(breaks=c(0:20),
  labels=c('Not Available','Debt Consolidation', 'Home Improvement', 
           'Business','Personal Loan','Student Use', 'Auto', 'Other', 
           'Baby&Adoption','Boat', 'Cosmetic Procedure', 'Engagement Ring', 
           'Green Loans', 'Household Expenses', 'Large Purchases', 
           'Medical/Dental', 'Motorcycle', 'RV', 'Taxes', 'Vacation', 
           'Wedding Loans'
 ))+
  stat_bin(aes(y=..count.., label=..count..), geom="text", vjust=-.5,size=3)+
  theme(axis.text.x = blue.bold.italic.16.text)+
  ggtitle("Listing category spread")
```

Most loans are for debt consolidation as seen in graph


```{r echo=FALSE, eval=FALSE,employeed_people_profile,cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
###Do employeed people default

ggplot(aes(x=LoanStatus,fill=EmploymentStatus),data=loan_data_csv)+
geom_histogram()+
theme(axis.text.x = blue.bold.italic.16.text)

ggplot(aes(x=EmploymentStatus,ymax=max(..count..)),
       data=subset(loan_data_csv,
                   (LoanStatus %in% c('Defaulted','ChargedOff'))))+
geom_histogram()+
theme(axis.text.x = blue.bold.italic.16.text)+
stat_bin(aes(y=..count.., label=..count..), geom="text", vjust=-.5,size=3)

#Lots of full time people default .Note 522 people who have not 
#specified employment status..
```






```{r echo=FALSE,eval=FALSE}
###Has the return in Prosper been good over the year for investors

#Plot->Lender yield over the years
#Comment->Lender yield has increased year on year except for 2009 
#We do not have the full data for 2014.

ggplot(aes(x=LenderYield),data=loan_data_csv)+
geom_freqpoly(binwidth=0.05)+
facet_wrap(~Loan_year)+
ggtitle("Lender yield per year")

#Clearly lender yield is going up year on year except for 2009
#2013 has been the best year 2014 data is only partially in

ggplot(aes(x=LenderYield,fill=factor(Loan_year)),
       data=subset(loan_data_csv,Loan_year %in% c(2011,2012,2013)))+
geom_histogram(binwidth=0.01)

table(loan_data_csv$Loan_year,loan_data_csv$LenderYield)

```



```{r echo=FALSE,eval=FALSE}
###Who is borrowing the loans 

#Plot->Borrower profile
#Comment->Most borrowers list other as occupation followed by professionals
#table(loan_data_csv$Occupation)
ggplot(aes(x=Occupation,ymax=max(..count..)),data=loan_data_csv)+geom_histogram()+
  theme(axis.text.x = blue.bold.italic.16.text)+
 stat_bin(aes(y=..count.., label=..count..), geom="text", vjust=-.5,size=4)+
  ggtitle("Borrower occupation")

library(dplyr)
#Group by and then see top 5
loan_data_by_occupation<-
  loan_data_csv %>%
  group_by(Occupation) %>%
  summarise( n=n()) %>%
  arrange(desc(n))
head(loan_data_by_occupation)

```



```{r echo=FALSE,eval=FALSE}
###How long have the borrowers been employed

#Plot->Borrower profile working history?
#Comment->Most loans go to full time employed but most people are employed for
#shorter durations


ggplot(aes(x=EmploymentStatus),data=loan_data_csv)+
geom_histogram()+
ggtitle("Employment Status spread")

#table(loan_data_csv$EmploymentStatusDuration)
summary(loan_data_csv$EmploymentStatusDuration)

#Positively skew Mean>Median People employed for shorter durations
ggplot(aes(x=EmploymentStatusDuration),data=loan_data_csv)+
geom_histogram(binwidth=10)+
ggtitle("borrower employment duration")



``` 



```{r echo=FALSE,eval=FALSE}
###How many credit lines do they have ?

table(loan_data_csv$CurrentCreditLines)
summary(loan_data_csv$CurrentCreditLines)
ggplot(aes(x=CurrentCreditLines),data=loan_data_csv)+
  geom_histogram(binwidth=2)+xlim(c(0,60))+
  ggtitle("Credit lines spread")
#Slight positive skew mean >median typically  have 10 credit lines
```

###Debt to income ratio of the borrowers

```{r echo=FALSE,eval=FALSE}


summary(loan_data_csv$DebtToIncomeRatio)
#table(loan_data_csv$DebtToIncomeRatio)
#Shows a positve skew mean is higher than median 
ggplot(aes(x=DebtToIncomeRatio),data=loan_data_csv)+
geom_histogram(fill='light blue')+
scale_x_continuous(breaks=seq(0,10.010,0.1))+
xlim(0,10.010)

```
```{r echo=FALSE}

#Fine tune for people with DI<1
library(ggplot2)
ggplot(aes(x=DebtToIncomeRatio), 
       data=subset(loan_data_csv,loan_data_csv$DebtToIncomeRatio<1)) + 
geom_histogram(binwidth=0.01,fill='light blue') +
ggtitle("Debt to Income Ratio<1 spread ")

```

Most people take on reasonable debt but there are few borrowers
who take on very large debt.
 
###Income range of the borrowers

```{r echo=FALSE}
#Plot->Income range distribution
#Comment->Largest group 25-50000
#table(loan_data_csv$IncomeRange)
summary(loan_data_csv$IncomeRange)
ggplot(aes(x=IncomeRange,ymax=max(..count..)),data=loan_data_csv)+geom_histogram()+
theme(axis.text.x = blue.bold.italic.16.text)+
stat_bin(aes(y=..count.., label=..count..), geom="text", vjust=-.5,size=4)+
ylim(0,33000)+
ggtitle("Income range spread")

```

Largest group of borrowers have an income range of 250000-50000 closely followed  by 50000-75000

```{r echo=FALSE,eval=FALSE}
#Monthly income distbrution There are some where monthy income is 0??? 
#Is that even a range? Check what kind of laons are given and how they service?

head(loan_data_csv[which(loan_data_csv$StatedMonthlyIncome==0),
                   c('StatedMonthlyIncome','IncomeRange')])
#Stated Monthly income =0 incomerange comes up as NotEmployed I dont think this
#is an appropriate income range

```



```{r echo=FALSE,eval=FALSE}
###Do people need recommendations to get a loan

table(loan_data_csv$Recommendations)
summary(loan_data_csv$Recommendations)
##Looks like recommendations are not required
```


```{r echo=FALSE,eval=FALSE}
###How many houseowners?

prop.table(table(loan_data_csv$IsBorrowerHomeowner))
```

###What is the lender yield like?

```{r echo=FALSE}
#Plot->LenderYield
#Comment->Slight positive skew maximum yield is close to 0.30
summary(loan_data_csv$LenderYield)
ggplot(aes(x=LenderYield),data=loan_data_csv)+
geom_histogram(binwidth=0.005,fill='light blue')  +
scale_x_continuous(breaks=seq(0, 0.5, 0.05)) + 
labs(title="The Lender Yield on the Loan")
```

Lender yield spread is positively skewed large percentage of loans yield close 
to 0.3%

# Univariate Analysis
###What is the structure of your dataset?
Data has 113937 observation with 81 variables
###What is/are the main feature(s) of interest in your dataset?
This data set has 81 variables so I chose a subset of the data as features to 
study.The features chosen are

*BorrowerState
*LoanOriginalAmount
*BorrowerAPR
*ProsperScore
*LenderYield
*CreditScoreRangeLower



###Did you create any new variables from existing variables in the dataset?
I created 6 variables Loan_year,Loan_month,Loan_closed_Date_month,
Loan_closed_Date_year,Credit_Type,TotalMonthlyDebt

### What other features in the dataset do you think will help support your
###investigation into your feature(s) of interest?
Other features I have used to futher support my investigation are .

*LoanOriginationDate
*ClosedDate
*Term
*BorrowerAPR
*ProsperScore
*ListingCategory
*Occupation
*EmploymentStatus
*EmploymentStatusDuration
*CurrentCreditLines
*DebtToIncomeRatio
*IncomeRange
*StatedMonthlyIncome
*Recommendations



### Of the features you investigated, were there any unusual distributions? 
###Did you perform any operations on the data to tidy, adjust, or change the
###form of the data? If so, why did you do this?
The plot of Loan volumes over the years shows that loans have abruptly fallen in
2009 which was unusual since it was growing year on year from 2005.


# Bivariate Plots Section
###Have lending criteria been stricter
```{r}

ggplot(aes(x=factor(Loan_year),fill=Credit_Type),data=loan_data_csv)+
geom_histogram()+
xlab("Loan_year")+
ggtitle("Borrower profile over the years")

```

From the graph we can see that credit requirements have become more stricter
from 2009 onwards loans are given to borrowers with atleast fair credit.

###Loan deliquency over time
```{r echo=FALSE}
library(dplyr)
defaults<-subset(loan_data_csv,(LoanStatus %in% c('Defaulted','ChargedOff')))
#table(defaults$Loan_year,defaults$LoanStatus)
#Number of defaulted
loan_status_default<-loan_data_csv %>%
  group_by(Loan_year) %>%
  summarise(CountTotal=n(), 
            DefaultTotal = sum(LoanStatus %in% c('Defaulted','ChargedOff'))) %>%
  mutate(percentile = DefaultTotal/CountTotal)

ggplot(aes(x=factor(Loan_year),y=percentile,group=1),data=loan_status_default)+
geom_line()+
geom_point()+
ggtitle("Percentage of defaults from 2005-2014")+
ylab('Percentage')+
xlab('Loan_year')


```

The plot shows default was very high in 2006 and then has fallen steadily after 
2009 to 2013 The next question would be why such a dramatic change 
i.e why have defaults fallen so much have lending standards improved


###Does loan's prosper score affect loan Interest rate?

```{r echo=FALSE}
cor.test(loan_data_csv$ProsperScore,loan_data_csv$BorrowerAPR)
```
Prosperscore and BorrowerAPR are strongly negatively corelated
```{r echo=FALSE}
ggplot(aes(y=BorrowerAPR,x=factor(ProsperScore)),
       data=subset(loan_data_csv,!is.na(ProsperScore)))+
geom_boxplot()+
ggtitle("Prosper score versus BorrowerAPR")+
xlab("Prosper score")

```

```{r echo=FALSE,eval=FALSE}

ggplot(aes(x=BorrowerAPR),data=subset(loan_data_csv,!is.na(ProsperScore)))+
geom_freqpoly(binwidth=0.05)+facet_wrap(~ProsperScore)+
ggtitle("BorrowerAPR spread")

#As prosperscore increases borrowerAPR decreases(distribution shifts left)
```

ProsperScore and BorrowerAPR are very negatively correlated
The graphs Prosper score vs BorrowerAPR and BorrowerAPR spread both indicate
that Interest rate fall for loans with high prosperscore and vice versa

###LoanOriginalAmount vs ProsperScore
```{r echo=FALSE}
cor.test(loan_data_csv$LoanOriginalAmount,loan_data_csv$ProsperScore)
```

Both loan original amount and prosperscore have a moderate corelation this
indicates Prosperscore increases these loans could potentially have a larger
loan amount

```{r echo=FALSE}

ggplot(aes(x=factor(ProsperScore),y=LoanOriginalAmount),data=subset(loan_data_csv,!is.na(ProsperScore)))+
geom_boxplot()+
xlab("Prosper score")+
ggtitle("Prosper score and loan amounts")

```

Plots indicate Loans with low prosperscores have low amounts and those with higher 
prosper scores can have higher loan amounts 



###Does credit score infulence borrower APR ?
```{r echo=FALSE}
cor.test(loan_data_csv$CreditScoreRangeLower,loan_data_csv$BorrowerAPR)
```

Both creditscore and borrower interest rate are strongly negatively correlated
indicating that borrowers with low credit score probably pay more interest and
borrowers with good credit score pay less interest

```{r echo=FALSE}

ggplot(aes(x=factor(CreditScoreRangeLower),y=BorrowerAPR),data=subset(loan_data_csv,!is.na(CreditScoreRangeLower)))+
geom_boxplot()+
xlab("Credit score")+
ggtitle("BorrowerAPR vs CreditScore")

```

The general trend is that as creditscore increases borrower APR decreases as evidenced by the graph.

###Income vs Loan Amount
```{r echo=FALSE}
cor.test(loan_data_csv$StatedMonthlyIncome,loan_data_csv$LoanOriginalAmount)

```

Both monthly income and loan amount are positively correlated this suggests that
people on higher incomes can potentially take on bigger loans
```{r echo=FALSE}

#People on larger Income get bigger loans
ggplot(aes(x=IncomeRange,y=LoanOriginalAmount),
       data=loan_data_csv)+
geom_boxplot()+
ggtitle("Income vs Loan Amount")

```

Plots show that people who are on larger incomes can take on larger loans

```{r echo=FALSE}

cor.test(loan_data_csv$Term,loan_data_csv$LoanOriginalAmount)
```

LoanOriginalAmount and loan term are positively correlated which implies larger 
loans are taken over a longer period

```{r echo=FALSE}
ggplot(aes(x=factor(Term),y=LoanOriginalAmount),data=loan_data_csv)+
geom_boxplot()+
ggtitle("Loan term vs Loan amount")+
xlab("Loan Term")

```

The above plot show that longer term loans are usually larger loans

```{r,echo=FALSE,eval=FALSE}
#Highter income and lower debtto income ratio this should be true
cor.test(loan_data_csv$StatedMonthlyIncome,loan_data_csv$DebtToIncomeRatio)
#negativey corelated i.e as monthly income increases debt to income decreases
#though Not as strongly negative as I imagined it would be 



```


```{r credit_Score_vs_loanamounts,cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE,eval=FALSE}
cor.test(loan_data_csv$CreditScoreRangeLower,loan_data_csv$LoanOriginalAmount)
Loan Amounts and credit scores are positively correlated this implies that 
bigger loan amounts could be granted to borrowers with better credit scores


ggplot(aes(x=CreditScoreRangeLower,y=LoanOriginalAmount),
       data=subset(loan_data_csv,!is.na(CreditScoreRangeLower)))+
geom_point(alpha=0.05,position=position_jitter(h=0))+
scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Prosper score', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
ggtitle("Credit score vs LoanAmount")+
xlab("Credit score ")+
ylab("Loan amount in $")

#From the graph we can conclude larger loans are given to borrowers with 
#good credit scores .

```


###More risk more reward? Lender yield vs Creditscore

```{r echo=FALSE}
cor.test(loan_data_csv$CreditScoreRangeLower,loan_data_csv$LenderYield)
```

CreditScore and Lenderyield negatively corelated implying as Credit Scores 
increases lender yield decreases and viceversa

```{r echo=FALSE}


ggplot(aes(x=factor(CreditScoreRangeLower),y=LenderYield),data=loan_data_csv)+
geom_boxplot()+
ggtitle("Lender yield vs Creditscore")+
xlab("Credit score")
```

As credit score increases lender yield trends down as evidenced by the plot

```{r echo=FALSE,eval=FALSE}
###Loan amount vs Lender yield

cor.test(loan_data_csv$LoanOriginalAmount,loan_data_csv$LenderYield)

ggplot(aes(y=LenderYield,x=LoanOriginalAmount),
       data=loan_data_csv)+
geom_point(alpha=0.01,position=position_jitter(h=0),color='orange')+
ggtitle('lender yield vs loan amount')

#Lender yield falls as loan amount increases as evidenced by the plot

```



###Lender yield and prosper score
```{r echo=FALSE}
cor.test(loan_data_csv$ProsperScore,loan_data_csv$LenderYield)

```

Lender yield and prosper score are highly negatively correlated
implying as prosper score increases lender yield decreases and vice versa

```{r echo=FALSE}
ggplot(aes(y=LenderYield,x=factor(ProsperScore)),
       data=subset(loan_data_csv,!is.na(ProsperScore)))+
geom_boxplot()+
ggtitle("Prosper score vs Lender yield")+
xlab("Prosper score")



```

The above plot reaffirms that for loans with high prosper scores the yield falls.

```{r echo=FALSE}
###Corelate monthly debt vs credit lines

loan_data_csv$TotalMonthlyDebt=loan_data_csv$DebtToIncomeRatio*
  loan_data_csv$StatedMonthlyIncome
```
```{r echo=FALSE,eval=FALSE}
cor.test(loan_data,_csv$TotalMonthlyDebt,loan_data_csv$CurrentCreditLines)

summary(loan_data_csv$TotalMonthlyDebt)
#Positively correlated as you have more credit lines you have more debt
ggplot(aes(y=TotalMonthlyDebt,x=CurrentCreditLines),data=loan_data_csv)+
geom_point(alpha=0.05)+
ylim(0,2000)

```

###Lender yield vs borrower apr
```{r echo=FALSE}
cor.test(loan_data_csv$LenderYield,loan_data_csv$BorrowerAPR)
ggplot(aes(x=BorrowerAPR,y=LenderYield),data=subset(loan_data_csv,!is.na(BorrowerAPR)))+
geom_point(alpha=0.05)+
geom_smooth(method='lm',color='red')+
ggtitle("Borrower APR vs Lender yield")

```

Both lender yield and Borrower APR are very highly corelated . The relationship is 
pretty linear

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the 
###investigation. How did the feature(s) of interest vary with other features
###in the dataset?

There is a strong negative relationship between prosperscore and borrowerapr 
meaning loans with higher scores have lower interest rate.

There is a very strong positive corelation between lender yield and borrowerAPR.
This implies that loans where lender yield increases have larger interest rates

There is a strong negative corelation between prosperscore and lender yield
This implies that loans with good prosper scores have lesser yield and viceversa

There is a strong negative correlation between creditscore and borrowerapr i.e
people with bigger creditscores get cheaper loans


LoanAmount and term of loans are positively correlated meaning Larger loans are 
taken over a longer period of time

There is a Negative correlation between credit scores and lender yield implying
more risk more reward


### Did you observe any interesting relationships between the other features
###(not the main feature(s) of interest)?
I found that the lending criteria of prosper has become more stringent .
Loans are given only to people with reasonably good Credit scores.
I also noticed default rates have fallen considerably over the years .

### What was the strongest relationship you found?
The most strongest relationship is between BorrowerAPR and lenderyield. 
LenderYield is high then borrowerAPR is high and viceversa. This could be true
since LenderYield is defined as interest rate less service fee i.e BorrowerAPR 
in some part determines lenderyield 

#Multivariate plots section




```{r lenderyield_vs_loanamount,cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE,eval=FALSE}
###Lender yield vs Loan amount 
ggplot(aes(y=LenderYield,x=LoanOriginalAmount,color=factor(ProsperScore)),
       data=subset(loan_data_csv,!is.na(ProsperScore)))+
geom_point(alpha=0.1)+
scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Prosper score', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
facet_wrap(~Loan_year)

#The above plot is interesting it shows that in 2009 lenders could get good 
#yields from low risk loans but of late lenders get high return only from
#high risk loans
```


###Prosper business how much have volumes grown in each state over the years

```{r loan_volumes,cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}

library(dplyr)
#Which state has taken out the most loans loan distribution
loan_data_by_state<-loan_data_csv %>%
                    group_by(BorrowerState,Loan_year) %>%
                    summarise(n=n())

ggplot(aes(x=Loan_year,y=n),
       data=subset(loan_data_by_state,
                   trimws(loan_data_by_state$BorrowerState)!=""))+
geom_line(color='blue')+
ylab("Loan volume ")+
facet_wrap(~BorrowerState)+
ggtitle("Loan volume spread ")
```

Some states have an abrupt distribution like IA,ME,ND after further
research I found that these states have disallowed prosper. CA seems to have 
the most loans followed by NY ,TX,GA,FL. RI,NV,SD dont have data for 2005-2008 
Prosper was introduced here after 2008. 





###Lenderyield versus DebtToIncomeRatio
```{r lenderyield_vs_debttoincome,cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}



#Where is the money?

ggplot(aes(y=LenderYield,x=DebtToIncomeRatio,color=factor(ProsperScore)),
       data=loan_data_csv)+
geom_point(na.rm=T)+
scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'ProsperScore', reverse = F,
    override.aes = list(alpha = 1, size = 2)))+
scale_x_continuous(breaks=seq(0,10,1))

#Scale x axis to look at majority of points 

ggplot(aes(y=LenderYield,x=DebtToIncomeRatio,color=factor(ProsperScore)),
       data=loan_data_csv)+
geom_point(na.rm=T,alpha=0.05)+
scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'ProsperScore', reverse = F,
    override.aes = list(alpha = 1, size = 2)))+ 
scale_x_continuous(breaks=seq(0,3,0.2),limit=c(0,1.6))+
scale_y_continuous(breaks=seq(0.1,0.35,0.05))
```

The money is in risky investments as evidenced by graph yield is high
where the DebtToIncomeRatio>1 .The prosper score of these loans is low.

```{r echo=FALSE,eval=FALSE}
cor.test(loan_data_csv$LenderYield,loan_data_csv$MonthlyLoanPayment)
cor.test(loan_data_csv$LoanOriginalAmount,loan_data_csv$MonthlyLoanPayment)
cor.test(loan_data_csv$StatedMonthlyIncome,loan_data_csv$LoanOriginalAmount)
summary(loan_data_csv$StatedMonthlyIncome)

```
###BorrowerAPR versus LoanAmount 
```{r echo=FALSE}


cor.test(loan_data_csv$BorrowerAPR,loan_data_csv$LoanOriginalAmount)
ggplot(aes(y=BorrowerAPR,x=LoanOriginalAmount,color=factor(Credit_Type)),
       data=subset(loan_data_csv,!is.na(Credit_Type)))+
geom_point(na.rm=T)+
scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Credit Type', reverse =F,
    override.aes = list(alpha = 1))) +
facet_wrap(~Loan_year)







```
Good credit scores are able to borrow at a low interest rate on larger loans
No particular relationship between loan amount and borrowerAPR

###Loan original amount vs Income
```{r echo=FALSE}
ggplot(aes(y=LoanOriginalAmount,x=StatedMonthlyIncome,color=EmploymentStatus),
       data=loan_data_csv)+
geom_point(na.rm=T,alpha=0.1)+
xlim(0,10000)+
scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'IncomeRange', reverse = F,
    override.aes = list(alpha = 1, size = 2)))


ggplot(aes(y=LoanOriginalAmount,x=StatedMonthlyIncome,color=IncomeRange),
       data=subset(loan_data_csv,!is.na(Credit_Type)))+
geom_point(na.rm = T,alpha=0.1)+
xlim(0,10000)+
scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'IncomeRange', reverse = F,
    override.aes = list(alpha = 1, size = 2)))+
facet_wrap(~Credit_Type)

```


From the above plots it can be concluded people who are employed and 
on a relatively highwage with good credit score take on higher debts. 


### OPTIONAL: Did you create any models with your dataset? Discuss the strengths
###and limitations of your model.
```{r  cache=TRUE, cache.path = 'cache/', fig.path='figure/'}


lr<-lm(LenderYield~BorrowerAPR+ProsperScore+
         CreditScoreRangeLower+
         DebtToIncomeRatio,data=loan_data_csv)

summary(lr)

#r^2 is 0.9876 .The linear model is very good at predicting lenderyield
#as evidenced by the R^2. The variables are significant hence I have included
#them all .Some of the independent variables have a high corelation among
#them so there could be a multicolinearity problem

library(car)
vif(lr)

#The VIFs are not too large so the model does not exhibit multi colinearity
```
A linear models is built predicting Lender yield using Borrower APR.
The linear model has an R^2 of 0.9876 which is pretty good. All independent 
variables are significant
```{r eval=FALSE,echo=FALSE}
#Linear model to predict Prosper score

lr<-lm(ProsperScore~BorrowerAPR+StatedMonthlyIncome+
         EmploymentStatusDuration+EmploymentStatus+
         OnTimeProsperPayments+CreditScoreRangeLower+
         DebtToIncomeRatio,
       data=loan_data_csv)



summary(lr)

#R^2 0.63. The linear model has very few variables but is not that great in
#predicting Prosperscore as evidenced by R^2
```
 


#Multivariate Analysis


### Talk about some of the relationships you observed in this part of the 
###investigation. Were there features that strengthened each other in terms of 
###looking at your feature(s) of interest?

I observed LoanAmounts versus monthly income using features IncomeRange,
EmploymentStatus ,Creditscore to understand the relationship further. After
looking at the plots I could conclude that people who are employed
with a good salary and reasonable credit scores take on larger loans.
I also found that lender yield increases in risky investments .
I also observed that Prosper's lending criteria has become more stricter to 
what it was a few years back.

### Were there any interesting or surprising interactions between features?
I observed loan volumes fell off in 2009 .After searching online I found that
the SEC had put a cease and desist order on Prosper in Nov 2008.It also appears
from the plots that Prosper have made their lending criteria more stringent 
from the time they started they seem to give loans only to people with good
credit history.


# Final Plots and Summary
###Prosper loan volumes by state
```{r echo=FALSE}
gg<-ggplot(aes(x=Loan_year,y=n),
    data=subset(loan_data_by_state,trimws(loan_data_by_state$BorrowerState)!=""))
gg<-gg+geom_line(color='blue')
gg<-gg+geom_point()
gg<-gg+facet_wrap(~BorrowerState, scales = "free_y")
gg<-gg+theme(axis.text.x = blue.bold.italic.16.text)+
     ggtitle("Loan volumes by year")
gg<-gg+xlab("Loan year")
gg<-gg+ylab("Number of loans")
gg
 ```

Loan volumes have increased drastically since 2005 with a dip in 2009.The plot 
also showed that prosper did not launch in all states simultaneously .In some 
states it started later like Rhode Island,nevada and south dakota and in some 
states its still not available like Maine, Iowa, and North Dakota.

###Tightening of Criteria for granting loans 
```{r echo=FALSE,loan_tightening, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}

loan_data_csv_year_credit<-
  loan_data_csv %>%
  filter(!is.na(CreditScoreRangeLower)) %>%
  group_by(Loan_year,CreditScoreRangeLower) %>%
  summarize(n=n())


loan_data_csv_by_year<-loan_data_csv %>%
                       group_by(Loan_year) %>%
                       summarize(total=n())



loan_data_combi<-merge(loan_data_csv_by_year,loan_data_csv_year_credit,
                       by='Loan_year')
loan_data_combi$percent<-loan_data_combi$n/loan_data_combi$total

ggplot(aes(x=CreditScoreRangeLower,y=percent,color=factor(Loan_year)),
       data=loan_data_combi)+
geom_line()+
geom_point()+
scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Loan_year', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
ggtitle("Borrower profile spread by year")+
xlab("Credit score ")+
ylab("Percentage of loans ")+
coord_cartesian(xlim=c(350,880))

```

Borower profile in terms of credit score lower has changed since 2006 to 2014 
In 2006 we did have some low  <500 loans and in 2014 there are no such loans 
all are above atleast 600 this could account for more defaults early on .

###Lender yield vs Borrower APR

```{r lenderyield_vs_borrowerapr,cache=TRUE, cache.path = 'cache/', fig.path='figure/',echo=FALSE}

ggplot(aes(y=LenderYield,x=BorrowerAPR,color=factor(ProsperScore)),
       data=subset(loan_data_csv,!is.na(BorrowerAPR)))+
  stat_smooth(method=lm,color='red')+
  geom_point(alpha=0.05)+
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Prosper score', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +
  coord_cartesian(xlim=c(0.05,0.42))+
  ggtitle("Lender yield vs BorrowerAPR")+
  xlab("Borrower interest rate in %")+
  ylab("Lender yield on loan in %")
  
```

The above plot shows that lender yield increases as borrowerAPR of loan 
increases.Notice that the relationship is linear as evidenced by the red line
The higher lender yield also corresponds to more riskier loans as 
evidenced by the color of the points .



#Reflection
This data set is pretty large with many different variables. My first difficulty
was understanding how the business peer to peer lending worked then I tried to 
understand what the various variables in the data set meant .Initially I choose 
far too many features then slowly I brought that down to a few main ones. 
Using EDA I then tried to explore their relationships. I wanted to understand 
why a lender/borrower would opt for p2p lending rather than go to a bank.
It would be useful if I could study what the investor return would be using p2p
and a brick and mortar bank or bonds or shares. 
Similarly for a borrower what the interest rate  would be for p2p and a 
standard bank.


I then tried to understand what drives the lender yield.The data inconculsively
shows that like all equities risky behaviour is rewarding 
I then tried to model what determines lender yield. 
My model has very few variables and a good R^2.I notice that
the independent variables are very well corelated so I try to check for multi
collinearity by calculating VIF(Variance Inflation Factors).These turn out to
be reasonable so I include all the independent variables in the model.It would
also be v interesting to model Prosper score.On what basis does Prosper allocate
Prosperscore to its loans .I did try modelling the same using a linear model but
my R^2 was not very good 0.63.

I can conclude that a platform like Prosper gives a good returns to investors 
the number of defaults over the years have fallen since granting of loans is 
screened more and only worthy borrowers are given loans. Since the P2P lending
space has become more competitive it would be interesting to see if the returns
that the investors are currently giving will continue or not.

